### 2018-06-26 latex/figurelegends.tex
First round of spell checking done

%%% TeX-master: "manuscript"

@@

\begin{figure}[ht]
  \centering
  \includegraphics{fig_1_pathway_overview}
  \caption{\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesized from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR). Nam is the main precursor and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases). For the recycling of Nam, two different pathways exist. The pathway found in yeast and many bacteria starts with the deamination of Nam by Nam deamidase (NADA). The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPRT). A third enzyme can consume Nam, the Nam N-methyltransferase (NNMT). For more details and abbreviations, see text.}
  \label{fig:pathway_overview}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
  \caption{\textbf{NNMT enables high NAD-consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates. NMMT enables higher steady-state NAD consumption flux despite reduced NAD concentrations.}
  \label{fig:NNMT_NAD_flux}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{fig_4_NamPRT_affinity_Nam}
  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis Menten constants ($K_M$) of NamPRT for Nam on steady-state NAD consumption flux and NAD concentration at different cell devision rates. In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell devision rates. In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables increasing NAD consumption flux and NAD concentration. NNMT furthermore makes both, consumption flux and concentration, relatively independent of cell division rates. Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell devision rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.}
  \label{fig:NamPRT_affinity_Nam}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreeasing $K_M$ of NNMT. The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
  \label{fig:optimal_substr_affinities}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{fig_6_unresolved_loop}
  \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A). Coloured circles indicate the enzymes present in the species besides NamPRT; blue: NNMT; black: NADA and NNMT; yellow: NADA. For a more comprehensive alignment, please see supplementary figure~S1. B) The visualisation of human NamPRT is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT using the model 2H3D as template \cite{Wang2006}. The inserted region is not resolved in crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red. C) The localisation of the FLAG-tagged mutant protein lacking the unresolved loop is not changed compared to wildtype human NamPRT. Both show a heterogenous nuclear cytosolic localisation in immunofluerescence images in HeLa S3 cells. But \textit{in vitro} measurements using recombinant protein show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E). Bars in D) and E) with different letters indicate significant difference as estimated with a T test assuming independent samples and significance at $p < 0.05$.}
  \label{fig:unresolved_loop}
\end{figure}

>>>>

%%% TeX-master: "manuscript"
% !TeX spellcheck=en_GB

@@

\begin{figure}[ht]
  \centering
  \includegraphics{fig_1_pathway_overview}
  \caption{\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR). Nam is the main precursor and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases). For the recycling of Nam, two different pathways exist. The pathway found in yeast and many bacteria starts with the deamination of Nam by Nam deamidase (NADA). The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPRT). A third enzyme can consume Nam, the Nam N-methyltransferase (NNMT). For more details and abbreviations, see text.}
  \label{fig:pathway_overview}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates. NNMT enables higher steady state NAD consumption flux despite reduced NAD concentrations.}
  \label{fig:NNMT_NAD_flux}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{fig_4_NamPRT_affinity_Nam}
  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_M$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates. In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates. In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables increasing NAD consumption flux and NAD concentration. NNMT furthermore makes both, consumption flux and concentration, relatively independent of cell division rates. Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.}
  \label{fig:NamPRT_affinity_Nam}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreasing $K_M$ of NNMT. The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
  \label{fig:optimal_substr_affinities}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{fig_6_unresolved_loop}
  \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A). Coloured circles indicate the enzymes present in the species besides NamPRT; blue: NNMT; black: NADA and NNMT; yellow: NADA. For a more comprehensive alignment, please see supplementary figure~S1. B) The visualisation of human NamPRT is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT using the model 2H3D as template \cite{Wang2006}. The inserted region is not resolved in crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red. C) The localisation of the FLAG-tagged mutant protein lacking the unresolved loop is not changed compared to wildtype human NamPRT. Both show a heterogeneous nuclear cytosolic localisation in immunofluorescence images in HeLa S3 cells. But \textit{in vitro} measurements using recombinant protein show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E). Bars in D) and E) with different letters indicate significant difference as estimated with a T test assuming independent samples and significance at $p < 0.05$.}
  \label{fig:unresolved_loop}
\end{figure}

### 2018-06-26 latex/introduction.tex
First round of spell checking done

%%% TeX-master: "manuscript"
% !TeX spellcheck=en_GB
\section{Introduction}

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism. Even though best known as cofactor for various redox-reactions , NAD is involved in a number of signalling processes that consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. These NAD+-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD+-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{?}. These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio. Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.\todo[author=Mathias B.]{Please add more references to the first paragraph. I think there are better references then my own review.}

The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated. However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions. The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours \cite{Liu2018}. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate. Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}. Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, (NR) \cite{Yoshino2018}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig. 1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD-synthesis pathway.  If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.  In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig. 1).

We here present a phylogenetic analysis of NAD-pathway in eukaryotes that. In contrast to our previous analysis, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotic pathway evolution. The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost. This was accompanied by a marked increase in the number of  NAD+-consuming signalling enzymes. To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam. Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development. Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.  Based on multiple sequence alignments,  we identified a sequence insertion in NamPRT in Deuterostomes  that is affecting the affinity of NamPRT,  undermining the predictions derived from our mathematical modelling approach

Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates.

>>>>

%%% TeX-master: "manuscript"
% !TeX spellcheck=en_GB

\section{Introduction}

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism. Even though best known as cofactor for various redox-reactions , NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD-dependent protein deacylation by sirtuins as well as the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{?}. These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio. Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.\todo[author=Mathias B.]{Please add more references to the first paragraph. I think there are better references then my own review.}

The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated. However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions. The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as 2 hours \cite{Liu2018}. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate. Imbalances in NAD homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}. Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, (NR) \cite{Yoshino2018}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig. 1), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig. 1).

We here present a phylogenetic analysis of NAD pathway in eukaryotes that. In contrast to our previous analysis, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotic pathway evolution. The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost. This was accompanied by a marked increase in the number of NAD-consuming signalling enzymes. To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam. Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development. Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete. Based on multiple sequence alignments, we identified a sequence insertion in NamPRT in Deuterostomes that is affecting the affinity of NamPRT, undermining the predictions derived from our mathematical modelling approach

Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD-dependent signalling mechanisms present in vertebrates.

### 2018-06-26 latex/methods.tex
First round of spell checking done

Kinetic parameters (substrate affinity (Km) and turnover rates (kcat), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions. Parameter values from mammals were used if available. For enzymes not present in mammals, values from yeast were used. The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1. For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}. Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively. All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}. The model will be available at the Biomodels database accession number xxx\todo[author=Ines]{Insert number}. Related figures were generated using gnuplot\todo[author=Ines]{Version?}.

@@

For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites. Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPT deletion mutant lacking amino acid residues 42-51 (NamPT-$\Delta$42-51). For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites. All cloned sequences were verified by DNA sequence analysis.

@@

The cells were harvested by centrifugation and resuspended in lysis buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 4\,mM dithiltheritol (DTT), 1\,mg/mL lysozyme, 1X Complete EDTA-free protease inhibitor cocktail (Roche)). After sonification, the lysate was centrifuged for 30\,min at 13000\,g, and the clear lysate was incubated with 2\,mL of Nickel-NTA resin (Qiagen). Non-specific protein binding was removed with washing buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 1\,mM DTT, 20\,mM imidazole). The protein was eluted with 2.5\,mL of elution buffer (20 mM\,Tris-HCl pH 8.0, 500\,mM NaCl, 300\,mM imidazole).

>>>>

Kinetic parameters (substrate affinity ($K_{M}$) and turnover rates ($k_{cat}$), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions. Parameter values from mammals were used if available. For enzymes not present in mammals, values from yeast were used. The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1. For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}. Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively. All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}. The model will be available at the Biomodels database accession number xxx\todo[author=Ines]{Insert number}. Related figures were generated using gnuplot\todo[author=Ines]{Version?}.

@@

For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites. Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51). For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites. All cloned sequences were verified by DNA sequence analysis.

@@

The cells were harvested by centrifugation and resuspended in lysis buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 4\,mM dithiothreitol (DTT), 1\,mg/mL lysozyme, 1X Complete EDTA-free protease inhibitor cocktail (Roche)). After sonification, the lysate was centrifuged for 30\,min at 13000\,g, and the clear lysate was incubated with 2\,mL of Nickel-NTA resin (Qiagen). Non-specific protein binding was removed with washing buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 1\,mM DTT, 20\,mM imidazole). The protein was eluted with 2.5\,mL of elution buffer (20 mM\,Tris-HCl pH 8.0, 500\,mM NaCl, 300\,mM imidazole).
